---
title: "Null Importance as a Testing Problem"
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
fig-width: 11
fig-height: 7
---

These are the dataset IDs used throughout.

| ID | Config name | Table 2 category | Paper example | Title |
|-----|----------------------|----------------------------------------|------------------------------|----------------------------------------------------------------------------------------------------------|
| D1 | linear_additive | Additive / Linear | -- | -- |
| D2 | quadratic | Additive / Quadratic | -- | -- |
| D3 | product_interaction | Interactions | -- | -- |
| D4 | parity | Interactions | Example 3.1 | Marginal null does not imply conditional null |
| D5 | dependent_features | Structural causal | -- | -- |
| D6 | confounding | Structural causal | -- | --|
| D7 | mediated_chains | Structural causal | Example 3.5, Example 4.1 | Causal importance without conditional statistical importance; Chain DAG counterexample (risk-based Shapley credits an indirect chain contribution) |
| D8 | redundant_pair | Structural causal | Example 3.3, Example 4.2 | Conditional statistical null without functional null; MDI fails to respect conditional null importance |
| D9 | bayes_incomplete | Additive / Quadratic | Example 3.4 | Failure of Bayes-completeness |
| D10 | heteroscedastic | Structural causal | Lemma 2 | Risk-based and conditional null relationship |

```{r setup, include=FALSE}
library(fs)
library(glue)
library(here)
library(tidyverse)
library(yaml)
source(here("case_studies", "src", "R", "visualize.R"))
source(here("case_studies", "src", "R", "error_rate_helpers.R"))
```

```{r}
base <- path(here("case_studies", "sweep_tabular"))
cfg <- read_yaml(base / "config.yaml")

evaluation_dir <- base / "results" / "evaluation"
fig_dir <- base / cfg$figures_dir
dir_create(fig_dir)
theme_set(theme_axiomatic(base_size = 12))
```

```{r}
response_type <- cfg$evaluation$response_type
ev <- load_evaluation(evaluation_dir) |> filter_response_type(response_type)
theoretical <- load_theoretical_table(
    base / "table1_theoretical.csv", cfg$theoretical_rows
)

save_fig <- function(plot, name, width = 11, height = 7) {
    ggsave(fig_dir / name, plot, width = width, height = height)
    plot
}
```

- $H_0(j, \text{notion})$ means that feature $j$ is null under that notion;
- a **type I error** calls a null feature relevant;
- a **type II error** misses a relevant feature.

 It's not entirely clear how to define the  rejection regions for each of the
 methods. So far, for minSHAP, we reject whenever we have a minSHAP score above
 0. For the other methods, we use the noise features that we deliberately
 introduced to define the decision boundary. However, in reality, these noise
 variables might not be available for calibration. The rates for those methods
 reflect a kind of best-case decision strategy.

## How do null importance criteria compare?

 We look at how many features are considered null versus non-null according to
 each notion of null importance, across all datasets considered in the
 benchmark.

```{r notion-separation}
#| fig-width: 6
#| fig-height: 5
notion_separation_panel(ev$truth) |>
    save_fig("fig_notion_separation.svg", width = 8, height = 6)
```

Notice that the power can depend a lot on sample size.

```{r confusion-conditional}
walk(cfg$sample_sizes, \(sample_size) {
    confusion_panel(ev$confusion, "conditional", sample_size = sample_size) |>
        save_fig(glue("fig_confusion_conditional_n{sample_size}.svg"), height = 9) |>
        print()
})
```

```{r confusion-causal}
walk(cfg$sample_sizes, \(sample_size) {
    confusion_panel(ev$confusion, "causal", sample_size = sample_size) |>
        save_fig(glue("fig_confusion_causal_n{sample_size}.svg"), height = 9) |>
        print()
})
```

## Mass placed on null features

Next we evaluate $\frac{\sum_{j \in \text{Null}}\left|\varphi_{j}\right|}{\sum_{j} \left|\varphi_{j}\right|}$. This can depend on the types of null importance considered.

```{r null-mass}
#| fig-height: 6
null_mass_panel(ev$null_mass, sample_size=500) |>
    save_fig("fig_null_mass_n_500.svg", width = 12, height = 6)

null_mass_panel(ev$null_mass, sample_size=5000) |>
    save_fig("fig_null_mass_n_5000.svg", width = 12, height = 6)
```

D7 nicely highlights the difference between SAGE and minSHAP.

```{r null-mass-d7}
ev$null_mass |>
    filter(
        dataset_id == "D7", notion == "risk",
        n == max(unlist(cfg$sample_sizes))
    ) |>
    summarise(
        across(c(null_magnitude, nonnull_magnitude), mean),
        .by = method
    ) |>
    mutate(null_mass = null_magnitude / (null_magnitude + nonnull_magnitude)) |>
    arrange(null_mass) |>
    knitr::kable(digits = 3)
```

## Dependence on the level

```{r table1}
#| fig-width: 14
#| fig-height: 6
walk(cfg$sample_sizes, \(sample_size) {
    table1_comparison_combined(ev$table1_by_n, sample_size = sample_size) |>
        save_fig(glue("fig_table1_empirical_n{sample_size}.svg"), width = 14, height = 6) |>
        print()
})
```

```{r table1-numbers}
ev$table1 |>
    select(
        method, notion, n_cells, controlled_fraction, powered_fraction,
        mean_fpr, mean_power, symbol, is_target_notion
    ) |>
    arrange(method, notion) |>
    knitr::kable(digits = 2)
```

## Theory against the sweep

```{r theory-vs-empirical}
theory_vs_empirical_panel(ev$table1, theoretical) |>
    save_fig("fig_theory_vs_empirical.svg", width = 9, height = 6)
```

## Summarizing paper $\iff$ code connections

This summarizes the data sets that we generate and compares it to the examples
from the paper draft.  We also summarize whether the empirical results are
consistent with the theory. There is some randomness involved in flagging a
feature as null/nonnull, and we heuristically gave some slack to tolerate
some departures on a fraction of datasets/features.

```{r counterexamples}
active_examples <- keep(
    cfg$counterexamples,
    \(entry) entry$dataset %in% ev$scores$dataset
)
ledger <- counterexample_ledger(ev$scores, active_examples)

ledger |>
    mutate(holds = if_else(holds, "yes", "no")) |>
    knitr::kable(digits = 2)
```

## Limitations

- We haven't evaluated `causal_shapley` so no method correctly flags causal nulls in D6 and D9
- We only considered D4 to be marginal nulls. In D2 and D3, the features are uncorrelated with the response, but not actually independent