---
title: "Sanity Checks for Synthetic Data Generation"
format: html
execute:
  echo: false
  warning: false
  message: false
fig-width: 10
fig-height: 4
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
source("../src/R/sanity_check_helpers.R")

# color palettes
palette_overall <- c(
    signal = "#1f77b4",
    null = "#d3d3d3",
    accent_orange = "#ff7f0e",
    accent_red = "#e74c3c"
)
palette_quantiles <- c("#d73027", "#fee090", "#1a9850")
sample_sizes <- c(50, 500, 5000)

# Apply theme globally with increased font sizes
theme_set(
    theme_classic() +
        theme(
            plot.title = element_text(size = 14, face = "bold", margin = margin(b = 8)),
            axis.title = element_text(size = 13),
            axis.text = element_text(size = 12),
            legend.position = "right",
            legend.title = element_text(size = 12),
            legend.text = element_text(size = 11),
            panel.grid.major.y = element_line(color = "#f0f0f0", size = 0.3),
            strip.text = element_text(size = 12, face = "bold")
        )
)
```

## Linear Additive

**DGP:** logit(Y) = β(x₁ + x₂ + ...)

```{r linear_additive_simple}
dat_500 <- read_dataset("linear_additive", 500)
features <- c("x1", "x2", "noise_1")
print(correlation_table(dat_500, features), digits = 3)
```

**Marginal relationship (n=500):**

```{r linear_additive_scatter, fig.width=14, fig.height=4.5}
compose_scatter_panels(dat_500,
    x_plain1 = "x1", x_binned = "x1", bin_feat = "x2",
    x_plain2 = "noise_1", palette_quantiles = palette_quantiles
)
```

**Across sample sizes:**

```{r linear_additive_facet, fig.width=14, fig.height=5}
plot_facet_across_n(
    "linear_additive", features,
    classify = ~ ifelse(.x == "noise_1", "Null", "Signal"),
    fill_values = c(Signal = palette_overall["signal"], Null = palette_overall["null"]),
    ns = sample_sizes
)
```

---

## XOR

**DGP:** logit(Y) = −γ·x₁·x₂, x_i ∈ {−1, +1}

```{r xor_simple}
dat_500 <- read_dataset("xor", 500)
features <- c("x1", "x2", "x3", "x4", "noise_1")
print(correlation_table(dat_500, features), digits = 3)
```

**Marginal relationship (n=500)**

```{r xor_scatter, fig.width=14, fig.height=4.5}
compose_scatter_panels(dat_500,
    x_plain1 = "x1", x_binned = "x1", bin_feat = "x2",
    x_plain2 = "noise_1", palette_quantiles = palette_quantiles
)
```

**Across sample sizes:**

```{r xor_facet, fig.width=14, fig.height=5}
plot_facet_across_n(
    "xor", features,
    classify = ~ ifelse(.x == "noise_1", "Null", "Signal"),
    fill_values = c(Signal = palette_overall["signal"], Null = palette_overall["null"]),
    ns = sample_sizes
)
```

---

## Product Interaction

**DGP:** logit(Y) = γ(x₁x₂ + x₃x₄ + ...)

```{r product_simple}
dat_500 <- read_dataset("product_interaction", 500)
features <- c("x1", "x2", "x3", "x4", "noise_1")
print(correlation_table(dat_500, features), digits = 3)
```

**Marginal relationship (n=500):**

```{r product_scatter, fig.width=14, fig.height=4.5}
compose_scatter_panels(dat_500,
    x_plain1 = "x1", x_binned = "x1", bin_feat = "x2",
    x_plain2 = "noise_1", palette_quantiles = palette_quantiles
)
```

**Across sample sizes:**

```{r product_facet, fig.width=14, fig.height=5}
plot_facet_across_n(
    "product_interaction", features,
    classify = ~ ifelse(.x == "noise_1", "Null", "Signal"),
    fill_values = c(Signal = palette_overall["signal"], Null = palette_overall["null"]),
    ns = sample_sizes
)
```

---

## Dependent Features

**DGP:** logit(Y) = γ(z₁ + z₂ + ...), with x₁ = z₁, x₂ = z₁ + ε (and similarly x₃, x₄)

In the plots below, x1, x3 are non-null anchors while x2, x4 are conditionally
null proxies.

```{r dependent_simple}
dat_500 <- read_dataset("dependent_features", 500)
features <- c("x1", "x2", "x3", "x4", "noise_1")
print(correlation_table(dat_500, features), digits = 3)
```

**Marginal relationship (n=500)**

```{r dependent_scatter, fig.width=14, fig.height=4.5}
compose_scatter_panels(dat_500,
    x_plain1 = "x1", x_binned = "x1", bin_feat = "x2",
    x_plain2 = "noise_1", palette_quantiles = palette_quantiles
)
```

**Across sample sizes:**

```{r dependent_facet, fig.width=14, fig.height=5}
plot_facet_across_n(
    "dependent_features", features,
    classify = ~ case_when(
        .x %in% c("x1", "x3") ~ "Anchor",
        .x %in% c("x2", "x4") ~ "Proxy (cond. null)",
        TRUE ~ "Null"
    ),
    fill_values = c(
        Anchor = palette_overall["signal"],
        `Proxy (cond. null)` = palette_overall["accent_orange"],
        Null = palette_overall["null"]
    ),
    ns = sample_sizes
)
```

---

## Confounding

**DGP:** logit(Y) = γ(z₁ + z₂ + ...), with x_j = z_j + ε (z_j unobserved)

The idea is that intervening on x_j would lead to no change in Y, even though
the variables are marginally and conditionally nonnull

```{r confounding_simple}
dat_500 <- read_dataset("confounding", 500)
features <- c("x1", "x2", "noise_1")
print(correlation_table(dat_500, features), digits = 3)
```

**Marginal relationship (n=500):**

```{r confounding_scatter, fig.width=14, fig.height=4.5}
compose_scatter_panels(dat_500,
    x_plain1 = "x1", x_binned = "x1", bin_feat = "x2",
    x_plain2 = "noise_1", palette_quantiles = palette_quantiles
)
```

**Across sample sizes:**

```{r confounding_facet, fig.width=14, fig.height=5}
plot_facet_across_n(
    "confounding", features,
    classify = ~ ifelse(.x == "noise_1", "Null", "Causally null\n(obs. assoc.)"),
    fill_values = c(
        Null = palette_overall["null"],
        `Causally null\n(obs. assoc.)` = palette_overall["signal"]
    ),
    ns = sample_sizes
)
```
