---
title: "Sanity Checks for Synthetic Data Generation"
format: html
execute:
  echo: false
  warning: false
  message: false
fig-width: 10
fig-height: 4
---

```{r setup, include=FALSE}
library(tidyverse)
library(patchwork)
source("../src/R/sanity_check_helpers.R")

# color palettes
palette_overall <- c(
    signal = "#1f77b4",
    null = "#d3d3d3",
    accent_orange = "#ff7f0e",
    accent_red = "#e74c3c"
)
palette_quantiles <- c("#d73027", "#fee090", "#1a9850")
active_seed <- 2026
options(sweep_tabular_seed = active_seed)
sample_sizes <- c(50, 500, 5000)

# Apply theme globally with increased font sizes
theme_set(
    theme_classic() +
        theme(
            plot.title = element_text(size = 14, face = "bold", margin = margin(b = 8)),
            axis.title = element_text(size = 13),
            axis.text = element_text(size = 12),
            legend.position = "right",
            legend.title = element_text(size = 12),
            legend.text = element_text(size = 11),
            panel.grid.major.y = element_line(color = "#f0f0f0", size = 0.3),
            strip.text = element_text(size = 12, face = "bold")
        )
)
```

Click the tabs to switch between regression and classification.

## Linear Additive

**DGP:** E[Y|x] = β(x₁ + x₂ + ...)

```{r linear_additive_setup, include=FALSE}
la_features <- c("x1", "x2", "noise_1")
la_classify <- ~ ifelse(.x == "noise_1", "Null", "Signal")
la_fill <- c(Signal = palette_overall["signal"], Null = palette_overall["null"])
```

::: {.panel-tabset}

### Classification

```{r linear_additive_classification_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("linear_additive", "classification", la_features)
```

```{r linear_additive_classification_facet, fig.width=14, fig.height=5}
plot_facet_across_n("linear_additive", "classification", la_features,
    classify = la_classify, fill_values = la_fill, ns = sample_sizes
)
```

### Regression

```{r linear_additive_regression_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("linear_additive", "regression", la_features)
```

```{r linear_additive_regression_facet, fig.width=14, fig.height=5}
plot_facet_across_n("linear_additive", "regression", la_features,
    classify = la_classify, fill_values = la_fill, ns = sample_sizes
)
```

:::

---

## XOR

**DGP:** E[Y|x] = −γ·prod(sign(x₁), sign(x₂), ...), x_i ∈ U[−1,1]

I realize that the "signal" variables are uniform while the nulls are gaussian.
I don't think it makes a difference conceptually, but we might want to revise to
enforce consistency.

```{r xor_setup, include=FALSE}
xor_features <- c("x1", "x2", "x3", "x4", "noise_1")
xor_classify <- ~ ifelse(.x == "noise_1", "Null", "Signal")
xor_fill <- c(Signal = palette_overall["signal"], Null = palette_overall["null"])
```

::: {.panel-tabset}

### Classification

```{r xor_classification_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("xor", "classification", xor_features)
```

```{r xor_classification_facet, fig.width=14, fig.height=5}
plot_facet_across_n("xor", "classification", xor_features,
    classify = xor_classify, fill_values = xor_fill, ns = sample_sizes
)
```

### Regression

```{r xor_regression_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("xor", "regression", xor_features)
```

```{r xor_regression_facet, fig.width=14, fig.height=5}
plot_facet_across_n("xor", "regression", xor_features,
    classify = xor_classify, fill_values = xor_fill, ns = sample_sizes
)
```

:::

---

## Product Interaction

**DGP:** E[Y|x] = γ(x₁x₂ + x₃x₄ + ...)

```{r product_setup, include=FALSE}
product_features <- c("x1", "x2", "x3", "x4", "noise_1")
product_classify <- ~ ifelse(.x == "noise_1", "Null", "Signal")
product_fill <- c(Signal = palette_overall["signal"], Null = palette_overall["null"])
```

::: {.panel-tabset}

### Classification

```{r product_classification_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("product_interaction", "classification", product_features)
```

```{r product_classification_facet, fig.width=14, fig.height=5}
plot_facet_across_n("product_interaction", "classification", product_features,
    classify = product_classify, fill_values = product_fill, ns = sample_sizes
)
```

### Regression

```{r product_regression_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("product_interaction", "regression", product_features)
```

```{r product_regression_facet, fig.width=14, fig.height=5}
plot_facet_across_n("product_interaction", "regression", product_features,
    classify = product_classify, fill_values = product_fill, ns = sample_sizes
)
```

:::

---

## Dependent Features

**DGP:** E[Y|z] = γ(z₁ + z₂ + ...), with x₁ = z₁, x₂ = z₁ + ε (and similarly x₃, x₄)

In the plots below, x1, x3 are non-null anchors while x2, x4 are conditionally
null proxies.

```{r dependent_setup, include=FALSE}
dependent_features_cols <- c("x1", "x2", "x3", "x4", "noise_1")
dependent_classify <- ~ case_when(
    .x %in% c("x1", "x3") ~ "Anchor",
    .x %in% c("x2", "x4") ~ "Proxy (cond. null)",
    TRUE ~ "Null"
)
dependent_fill <- c(
    Anchor = palette_overall["signal"],
    `Proxy (cond. null)` = palette_overall["accent_orange"],
    Null = palette_overall["null"]
)
```

::: {.panel-tabset}

### Classification

```{r dependent_classification_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("dependent_features", "classification", dependent_features_cols)
```

```{r dependent_classification_facet, fig.width=14, fig.height=5}
plot_facet_across_n("dependent_features", "classification", dependent_features_cols,
    classify = dependent_classify, fill_values = dependent_fill, ns = sample_sizes
)
```

### Regression

```{r dependent_regression_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("dependent_features", "regression", dependent_features_cols)
```

```{r dependent_regression_facet, fig.width=14, fig.height=5}
plot_facet_across_n("dependent_features", "regression", dependent_features_cols,
    classify = dependent_classify, fill_values = dependent_fill, ns = sample_sizes
)
```

:::

---

## Quadratic

**DGP:** E[Y|x] = γ·∑(x₁² − 1, …, xₙ² − 1)

Each signal feature has zero marginal covariance with the response because the
response depends only on xⱼ². However, the features are relevant from
functional, conditional, and causal perspectives.

```{r quadratic_setup, include=FALSE}
quadratic_features <- c("x1", "x2", "x3", "x4", "noise_1")
quadratic_classify <- ~ ifelse(.x == "noise_1", "Null", "Signal")
quadratic_fill <- c(Signal = palette_overall["signal"], Null = palette_overall["null"])
```

::: {.panel-tabset}

### Classification

```{r quadratic_classification_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("quadratic", "classification", quadratic_features)
```

```{r quadratic_classification_facet, fig.width=14, fig.height=5}
plot_facet_across_n("quadratic", "classification", quadratic_features,
    classify = quadratic_classify, fill_values = quadratic_fill, ns = sample_sizes
)
```

### Regression

```{r quadratic_regression_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("quadratic", "regression", quadratic_features)
```

```{r quadratic_regression_facet, fig.width=14, fig.height=5}
plot_facet_across_n("quadratic", "regression", quadratic_features,
    classify = quadratic_classify, fill_values = quadratic_fill, ns = sample_sizes
)
```

:::

---

## Confounding

**DGP:** E[Y|z] = γ(z₁ + z₂ + ...), with x_j = z_j + ε (z_j unobserved)

The idea is that intervening on x_j would lead to no change in Y, even though
the variables are marginally and conditionally nonnull

```{r confounding_setup, include=FALSE}
confounding_features <- c("x1", "x2", "noise_1")
confounding_classify <- ~ ifelse(.x == "noise_1", "Null", "Causally null\n(obs. assoc.)")
confounding_fill <- c(
    Null = palette_overall["null"],
    `Causally null\n(obs. assoc.)` = palette_overall["signal"]
)
```

::: {.panel-tabset}

### Classification

```{r confounding_classification_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("confounding", "classification", confounding_features)
```

```{r confounding_classification_facet, fig.width=14, fig.height=5}
plot_facet_across_n("confounding", "classification", confounding_features,
    classify = confounding_classify, fill_values = confounding_fill, ns = sample_sizes
)
```

### Regression

```{r confounding_regression_scatter, fig.width=14, fig.height=4.5}
render_correlation_and_scatter("confounding", "regression", confounding_features)
```

```{r confounding_regression_facet, fig.width=14, fig.height=5}
plot_facet_across_n("confounding", "regression", confounding_features,
    classify = confounding_classify, fill_values = confounding_fill, ns = sample_sizes
)
```

:::
