---
title: "Method Comparison on Synthetic Data"
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
fig-width: 10
fig-height: 6
---

```{r setup, include=FALSE}
library(fs)
library(ggbump)
library(ggrepel)
library(glue)
library(here)
library(patchwork)
library(scico)
library(styler)
library(tidytext)
library(tidyverse)
library(yaml)

source(here("case_studies", "src", "R", "visualize.R"))
source(here("case_studies", "src", "R", "importance_visualize_helpers.R"))

base <- path(here("case_studies", "sweep_tabular"))
cfg <- read_yaml(base / "config.yaml")

methods <- names(cfg$methods)[unlist(cfg$methods)]
datasets <- names(cfg$datasets)
sample_sizes <- cfg$sample_sizes
seeds <- cfg$seeds
results_dir <- base / "results"
fig_dir <- base / cfg$figures_dir
dir_create(fig_dir)

theme_set(theme_axiomatic(base_size = 12))
null_colors <- null_type_colors()

mediated_focus <- list(n = 5000, response_type = "regression", seed = 2026)
```

This notebook visualizes the output of

```
python case_studies/sweep_tabular/sweep.py
```

We consider 7 explanation methods × 5 datasets × 10 seeds three ways,

- MFA: map of every (sample size, seed) importance table, the features are
  samples, blocked so no one run dominates.
- Bump plot: each method's rank trajectory across features.
- Zoom: marginal vs. conditional importance on one correlated pair.

## Multiple Factor Analysis

Blocking by sample size keeps any one sample size from dominating the
projection. Panels are sorted by the variance explained by the first axis, which
separates signal from noise features.  For `dependent_features`, where true
"anchors" are correlated with irrelevant "proxies", correlation clearly fails.
Knockoffs and correlation both struggle on `product_interaction` and `parity`,
which isn't surprising given the nonlinear structure.

```{r mfa-compute}
mfa_by_dataset <- map(datasets, \(ds) {
    mfa_by_sample_size(
        ds,
        ns = sample_sizes,
        response_type = "regression",
        seeds = seeds,
        methods = methods,
        results_dir = results_dir
    )
})
names(mfa_by_dataset) <- datasets

# Order datasets by decreasing Dim1 variance explained.
dim1_var <- map_dbl(mfa_by_dataset, ~ .x$var_explained[1])
datasets_by_dim1 <- names(sort(dim1_var, decreasing = TRUE))

panel_labels <- imap_chr(mfa_by_dataset, \(res, ds) {
    glue(
        "{ds}\n({round(res$var_explained[1], 1)}%, ",
        "{round(res$var_explained[2], 1)}%)"
    )
})
panel_levels <- panel_labels[datasets_by_dim1]

mfa_df <- imap_dfr(mfa_by_dataset, \(res, ds) {
    res$coords |> mutate(dataset = ds, panel_label = panel_labels[[ds]])
}) |>
    left_join(null_type_table(), by = c("dataset", "feature")) |>
    mutate(panel_label = factor(panel_label, levels = panel_levels))

partial_df <- imap_dfr(mfa_by_dataset, \(res, ds) {
    res$partial |> mutate(dataset = ds, panel_label = panel_labels[[ds]])
}) |>
    left_join(null_type_table(), by = c("dataset", "feature")) |>
    mutate(panel_label = factor(panel_label, levels = panel_levels))

loadings_df <- imap_dfr(mfa_by_dataset, \(res, ds) {
    res$loadings |>
        mutate(dataset = ds, panel_label = panel_labels[[ds]])
}) |>
    mutate(panel_label = factor(panel_label, levels = panel_levels))
```

```{r fig-mfa-scores, fig.width=10, fig.height=6}
p_mfa_scores <- ggplot(mfa_df, aes(Dim.1, Dim.2)) +
    geom_hline(yintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_vline(xintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_point(
        data = partial_df, aes(color = null_type),
        size = 1, alpha = 0.35, show.legend = FALSE
    ) +
    geom_point(aes(color = null_type), size = 3) +
    geom_text_repel(
        data = filter(mfa_df, !str_detect(feature, "noise")),
        aes(color = null_type, label = feature),
        size = 3, show.legend = FALSE, max.overlaps = 20
    ) +
    facet_wrap(~panel_label, scales = "free", nrow = 2) +
    scale_color_manual(values = null_colors, name = "Null type") +
    labs(
        title = "MFA scores",
        x = "Dimension 1", y = "Dimension 2"
    )
p_mfa_scores
ggsave(
    path(fig_dir, "fig_mfa_scores.png"), p_mfa_scores, width = 10, height = 6,
    dpi = 200
)
```

```{r fig-mfa-loadings, fig.width=10, fig.height=5}
loadings_long <- loadings_df |>
    pivot_longer(c(Dim.1, Dim.2), names_to = "dim", values_to = "loading") |>
    mutate(
        dataset = factor(dataset, levels = rev(datasets_by_dim1)),
        dim = recode(dim, Dim.1 = "Dimension 1", Dim.2 = "Dimension 2"),
        sign = if_else(loading >= 0, "positive", "negative")
    )

# Cluster methods independently within each MFA dimension.  Each method is
# represented by its loading profile across datasets, so nearby methods have
# similar profiles in the corresponding facet.
method_orders <- loadings_long |>
    select(dim, method, dataset, loading) |>
    pivot_wider(names_from = dataset, values_from = loading) |>
    group_nest(dim) |>
    mutate(method_order = map(data, function(d) {
        method_matrix <- d |>
            select(-method) |>
            as.matrix()
        rownames(method_matrix) <- d$method

        hc <- hclust(dist(method_matrix), method = "ward.D2")
        tibble(
            method = rownames(method_matrix)[hc$order],
            cluster_order = seq_along(hc$order)
        )
    })) |>
    select(dim, method_order) |>
    unnest(method_order)

loadings_long <- loadings_long |>
    left_join(method_orders, by = c("dim", "method")) |>
    mutate(method_ordered = reorder_within(
        method, cluster_order, dim, sep = "___"
    ))

p_mfa_loadings <- ggplot(loadings_long, aes(method_ordered, dataset)) +
    geom_point(
        aes(size = abs(loading), fill = sign),
        shape = 22, color = axiom_palette$ink, stroke = 0.4
    ) +
    facet_wrap(~dim, nrow = 1, scales = "free_x") +
    scale_x_reordered() +
    scale_size_area(max_size = 10, limits = c(0, 1), name = "|Loading|") +
    scale_fill_manual(
        values = c(positive = axiom_palette$ink, negative = "white"),
        name = "Sign"
    ) +
    labs(
        title = "MFA loadings",
        x = "Method", y = NULL
    ) +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        panel.background = element_rect(fill = "#cecdcd", color = NA),
        panel.grid.major = element_blank(),
        strip.text = element_text(size=10)
    )
p_mfa_loadings
ggsave(
    path(fig_dir, "fig_mfa_loadings.png"), p_mfa_loadings,
    width = 10, height = 5,
    dpi = 200
)
```

## Marginal vs. conditional null importance

Here is a "zoomed in" comparison between marginal and conditional null
importance. In the `dependent_features` data, we have correlated pairs where
only one ("anchors") are actually in the model. Knockoffs successfully
distinguishes these variable types, but marginal correlation can't.

```{r divergence-compute}
divergence_configs <- expand_grid(
    seed = seeds, n = sample_sizes, response_type = cfg$response_types
)

divergence_df <- pmap_dfr(divergence_configs, \(seed, n, response_type) {
    corr <- load_method_importance(
        "dependent_features", n, response_type, seed, "correlation",
        results_dir
    ) |>
        rename(correlation = importance)
    kn <- load_method_importance(
        "dependent_features", n, response_type, seed, "knockoffs",
        results_dir
    ) |>
        rename(knockoffs = importance)

    corr |>
        select(feature, correlation) |>
        left_join(select(kn, feature, knockoffs), by = "feature") |>
        filter(feature %in% c("x1", "x2")) |>
        mutate(seed = seed, n = n, response_type = response_type)
})
```

```{r fig-divergence, fig.width=7.5, fig.height=3}
feature_colors <- c(x1 = "#31abc0", x2 = "#ab4dca")
feature_labels <- c(x1 = "x1 (signal)", x2 = "x2 (correlated proxy)")

# Precompute per-group means to overlay on the plot
divergence_means <- divergence_df |>
    group_by(feature, response_type, n) |>
    summarize(
        correlation = mean(correlation), knockoffs = mean(knockoffs),
        .groups = "drop"
    )

p_div <- ggplot(
    divergence_df,
    aes(correlation, knockoffs, color = feature, shape = response_type)
) +
    geom_hline(yintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_vline(xintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_point(size = 2, alpha = 0.4) +
    geom_point(data = divergence_means, size = 3, alpha = 0.8) +
    facet_wrap(~n, labeller = labeller(n = ~ glue("n = {.x}"))) +
    scale_color_manual(
        values = feature_colors, labels = feature_labels, name = NULL
    ) +
    scale_shape_manual(values = c(classification = 16, regression = 17)) +
    labs(
        title = glue("Correlation vs. knockoffs on dependent_features"),
        x = "Correlation importance", y = "Knockoffs importance",
        shape = "Response type"
    )

p_div
```

# minSHAP mediated-chain diagnostic

The synthetic mediated-chain case study has two chains,
$x_1 \to x_2 \to x_3 \to Y$ and
$x_4 \to x_5 \to x_6 \to Y$. Only $x_3$ and $x_6$ are conditionally related
to the response after all features are observed. The plots below use the exact
sampled orderings and predecessor coalitions that produced the reported
minSHAP and SAGE results. They do not enumerate a separate coalition game.

```{r load-mediated-chain-contributions}
mediated_stem <- with(
    mediated_focus,
    glue("mediated_chains_{n}_{response_type}_{seed}")
)
mediated <- load_mediated_contributions(mediated_stem, results_dir)
mediated_contributions <- mediated$contributions
mediated_membership <- mediated$membership
```

## Exact sampled conditional contributions

```{r fig-mediated-contribution-focus, fig.width=15, fig.height=18}
focus <- mediated_focus_data(mediated_membership, mediated_contributions, FEATURE_ORDER)
p_mediated_focus <- mediated_membership_plot(
    focus$membership, focus$outline
) | mediated_bars_plot(focus$bars)
p_mediated_focus

ggsave(
    path(fig_dir, "fig_mediated_chain_contribution_focus.png"),
    p_mediated_focus, width = 15, height = 18,
    dpi = 200
)
```

Panel A shows the correlation/knockoffs divergence, while panel B shows the
sampled minSHAP contributions.

```{r fig-divergence-and-minshap, fig.width=10, fig.height=9}
mediation_highlight <- c("noise_1", "x1", "x3")
highlight <- mediated_focus_data(
    mediated_membership, mediated_contributions, mediation_highlight
)

p_minshap <- mediated_membership_plot(
    highlight$membership, highlight$outline,
    title = "Coalitions"
) + labs(tag = "B")

p_combined <- (p_div + labs(tag = "A")) /
    (p_minshap | mediated_bars_plot(highlight$bars)) +
    plot_layout(heights = c(1, 1.3)) &
    theme(plot.tag = element_text(face = "bold", size = 14))
p_combined

ggsave(
    path(fig_dir, "fig_knockoffs_and_minshap.png"), p_combined,
    width = 10, height = 9,
    dpi = 200
)
```
