---
title: "Method Comparison on Synthetic Data"
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
fig-width: 10
fig-height: 6
---

```{r setup, include=FALSE}
library(fs)
library(ggbump)
library(ggrepel)
library(glue)
library(here)
library(patchwork)
library(scico)
library(styler)
library(tidytext)
library(tidyverse)
library(yaml)

source(here("case_studies", "src", "R", "visualize.R"))
source(here("case_studies", "src", "R", "importance_visualize_helpers.R"))

base <- path(here("case_studies", "sweep_tabular"))
cfg <- read_yaml(base / "config.yaml")

methods <- names(cfg$methods)[unlist(cfg$methods)]
datasets <- names(cfg$datasets)
sample_sizes <- cfg$sample_sizes
seeds <- cfg$seeds
results_dir <- base / "results"
fig_dir <- base / cfg$figures_dir
dir_create(fig_dir)

theme_set(theme_axiomatic(base_size = 12))
null_colors <- null_type_colors()

mediated_focus <- list(n = 5000, response_type = "regression", seed = 2026)
```

This notebook visualizes the output of

```
python case_studies/sweep_tabular/sweep.py
```

We consider 7 explanation methods × 5 datasets × 10 seeds three ways,

- MFA: map of every (sample size, seed) importance table, the features are
  samples, blocked so no one run dominates.
- Bump plot: each method's rank trajectory across features.
- Zoom: marginal vs. conditional importance on one correlated pair.

## Multiple Factor Analysis

Blocking by sample size keeps any one sample size from dominating the
projection. Panels are sorted by the variance explained by the first axis, which
separates signal from noise features.  For `dependent_features`, where true
"anchors" are correlated with irrelevant "proxies", correlation clearly fails.
Knockoffs and correlation both struggle on `product_interaction` and `parity`,
which isn't surprising given the nonlinear structure.

```{r mfa-compute}
mfa_by_dataset <- map(datasets, \(ds) {
    mfa_by_sample_size(
        ds,
        ns = sample_sizes,
        response_type = "classification",
        seeds = seeds,
        methods = methods,
        results_dir = results_dir
    )
})
names(mfa_by_dataset) <- datasets

# Order datasets by decreasing Dim1 variance explained.
dim1_var <- map_dbl(mfa_by_dataset, ~ .x$var_explained[1])
datasets_by_dim1 <- names(sort(dim1_var, decreasing = TRUE))

panel_labels <- imap_chr(mfa_by_dataset, \(res, ds) {
    glue(
        "{ds}\n(Dim1 {round(res$var_explained[1], 1)}%, ",
        "Dim2 {round(res$var_explained[2], 1)}%)"
    )
})
panel_levels <- panel_labels[datasets_by_dim1]

mfa_df <- imap_dfr(mfa_by_dataset, \(res, ds) {
    res$coords |> mutate(dataset = ds, panel_label = panel_labels[[ds]])
}) |>
    left_join(null_type_table(), by = c("dataset", "feature")) |>
    mutate(panel_label = factor(panel_label, levels = panel_levels))

partial_df <- imap_dfr(mfa_by_dataset, \(res, ds) {
    res$partial |> mutate(dataset = ds, panel_label = panel_labels[[ds]])
}) |>
    left_join(null_type_table(), by = c("dataset", "feature")) |>
    mutate(panel_label = factor(panel_label, levels = panel_levels))

loadings_df <- imap_dfr(mfa_by_dataset, \(res, ds) {
    res$loadings |>
        mutate(dataset = ds, panel_label = panel_labels[[ds]])
}) |>
    mutate(panel_label = factor(panel_label, levels = panel_levels))
```

```{r fig-mfa-scores, fig.width=10, fig.height=6}
p_mfa_scores <- ggplot(mfa_df, aes(Dim.1, Dim.2)) +
    geom_hline(yintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_vline(xintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_point(
        data = partial_df, aes(color = null_type),
        size = 1, alpha = 0.35, show.legend = FALSE
    ) +
    geom_point(aes(color = null_type), size = 3) +
    geom_text_repel(
        aes(color = null_type, label = feature),
        size = 3, show.legend = FALSE, max.overlaps = 20
    ) +
    facet_wrap(~panel_label, scales = "free", nrow = 2) +
    scale_color_manual(values = null_colors, name = "Null type") +
    labs(
        title = "Feature Importance MFA: scores",
        x = "Dimension 1", y = "Dimension 2"
    )
p_mfa_scores
ggsave(
    path(fig_dir, "fig_mfa_scores.svg"), p_mfa_scores, width = 10, height = 6
)
```

```{r fig-mfa-loadings, fig.width=10, fig.height=5}
loadings_long <- loadings_df |>
    pivot_longer(c(Dim.1, Dim.2), names_to = "dim", values_to = "loading") |>
    mutate(
        dataset = factor(dataset, levels = rev(datasets_by_dim1)),
        dim = recode(dim, Dim.1 = "Dimension 1", Dim.2 = "Dimension 2"),
        sign = if_else(loading >= 0, "positive", "negative")
    )

p_mfa_loadings <- ggplot(loadings_long, aes(method, dataset)) +
    geom_point(
        aes(size = abs(loading), fill = sign),
        shape = 22, color = axiom_palette$ink, stroke = 0.4
    ) +
    facet_wrap(~dim, nrow = 1) +
    scale_size_area(max_size = 10, limits = c(0, 1), name = "|Loading|") +
    scale_fill_manual(
        values = c(positive = axiom_palette$ink, negative = "white"),
        name = "Sign"
    ) +
    labs(
        title = "Feature Importance MFA: loadings",
        x = "Method", y = NULL
    ) +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1),
        panel.background = element_rect(fill = "#cecdcd", color = NA),
        panel.grid.major = element_blank()
    )
p_mfa_loadings
ggsave(
    path(fig_dir, "fig_mfa_loadings.svg"), p_mfa_loadings,
    width = 10, height = 5
)
```

## Variable importance rankings

These are rankings at $n = 500$ for the classification tasks, with panels
ordered by variance in rankings. This parallels the MFA, with correlation and
knockoffs struggling with nonlinear data and correlation is failing for
dependent features. There are also some ties.
In real data this would be unusual, but we are explaining the simulated function
itself (not a fitted model), so the feature importances are actually equal —
e.g., for linear additive, noise features aren't in the model, so predictions
don't change when permuted.

```{r fig-bump, fig.width=15, fig.height=8}
n_features <- length(FEATURE_ORDER)
strip_y <- n_features + 0.9
method_shapes <- setNames(
    c(15, 16, 17, 18, 3, 4, 8, 11)[seq_along(methods)], methods
)

ranks_by_dataset <- map(datasets, \(ds) {
    map_dfr(seeds, \(s) {
        method_ranks(
            ds, n = 500, response_type = "classification", seed = s,
            methods = methods, results_dir = results_dir
        ) |>
            mutate(seed = s)
    })
})
names(ranks_by_dataset) <- datasets

# Order datasets by variance in their feature rankings
datasets_by_disagreement <- names(sort(map_dbl(ranks_by_dataset, \(df) {
    df |>
        group_by(feature) |>
        summarize(v = var(rank)) |>
        pull(v) |>
        mean(na.rm = TRUE)
}), decreasing = TRUE))

# Define legends that will appear in the last panel of the patchwork grid
null_legend_panel <- tibble(null_type = names(null_colors)) |>
    mutate(
        null_type = fct_rev(factor(null_type, levels = names(null_colors)))
    ) |>
    ggplot(aes(x = 1, y = null_type, fill = null_type)) +
    geom_tile(width = 0.5, height = 0.7, show.legend = FALSE) +
    geom_text(aes(label = null_type), x = 1.35, hjust = 0, size = 4) +
    scale_fill_manual(values = null_colors) +
    xlim(0.6, 3) +
    labs(title = "Null type") +
    theme_void() +
    theme(
        plot.title = element_text(hjust = 0, size = rel(1.1)),
        plot.background = element_rect(fill = axiom_palette$bg, color = NA)
    )

method_legend_panel <- tibble(
    method = fct_rev(factor(methods, levels = methods))
) |>
    ggplot(aes(x = 1, y = method, shape = method)) +
    geom_point(
        size = 2.8, stroke = 1, color = axiom_palette$ink, show.legend = FALSE
    ) +
    geom_text(aes(label = method), x = 1.35, hjust = 0, size = 4) +
    scale_shape_manual(values = method_shapes) +
    xlim(0.6, 3) +
    labs(title = "Method") +
    theme_void() +
    theme(
        plot.title = element_text(hjust = 0, size = rel(1.1)),
        plot.background = element_rect(fill = axiom_palette$bg, color = NA)
    )

legend_panel <- (null_legend_panel | method_legend_panel)
bump_panels <- map(datasets_by_disagreement, bump_panel)
p_bump <- wrap_plots(
    c(bump_panels, list(legend_panel)), ncol = 2, guides = "collect"
) +
    plot_annotation(title = glue("Importance Rankings (n=500, classification)"))
p_bump
ggsave(path(fig_dir, "fig_rank_bump.svg"), p_bump, width = 13, height = 7)
```

## Marginal vs. conditional null importance

Here is a "zoomed in" comparison between marginal and conditional null
importance. In the `dependent_features` data, we have correlated pairs where
only one ("anchors") are actually in the model. Knockoffs successfully
distinguishes these variable types, but marginal correlation can't.

```{r divergence-compute}
divergence_configs <- expand_grid(
    seed = seeds, n = sample_sizes, response_type = cfg$response_types
)

divergence_df <- pmap_dfr(divergence_configs, \(seed, n, response_type) {
    corr <- load_method_importance(
        "dependent_features", n, response_type, seed, "correlation",
        results_dir
    ) |>
        rename(correlation = importance)
    kn <- load_method_importance(
        "dependent_features", n, response_type, seed, "knockoffs",
        results_dir
    ) |>
        rename(knockoffs = importance)

    corr |>
        select(feature, correlation) |>
        left_join(select(kn, feature, knockoffs), by = "feature") |>
        filter(feature %in% c("x1", "x2")) |>
        mutate(seed = seed, n = n, response_type = response_type)
})
```

```{r fig-divergence, fig.width=7.5, fig.height=3}
feature_colors <- c(x1 = "#31abc0", x2 = "#ab4dca")
feature_labels <- c(x1 = "x1 (signal)", x2 = "x2 (correlated proxy)")

# Precompute per-group means to overlay on the plot
divergence_means <- divergence_df |>
    group_by(feature, response_type, n) |>
    summarize(
        correlation = mean(correlation), knockoffs = mean(knockoffs),
        .groups = "drop"
    )

p_div <- ggplot(
    divergence_df,
    aes(correlation, knockoffs, color = feature, shape = response_type)
) +
    geom_hline(yintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_vline(xintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_point(size = 2, alpha = 0.4) +
    geom_point(data = divergence_means, size = 3, alpha = 0.8) +
    facet_wrap(~n, labeller = labeller(n = ~ glue("n = {.x}"))) +
    scale_color_manual(
        values = feature_colors, labels = feature_labels, name = NULL
    ) +
    scale_shape_manual(values = c(classification = 16, regression = 17)) +
    labs(
        title = glue("Correlation vs. knockoffs on dependent_features"),
        x = "Correlation importance", y = "Knockoffs importance",
        shape = "Response type"
    )

p_div
ggsave(
    path(fig_dir, "fig_divergence_dependent_features.svg"), p_div,
    width = 7.5, height = 3
)
```

```{r divergence-compute-corr}
divergence_df_corr <- pmap_dfr(divergence_configs, \(seed, n, response_type) {
    corr <- load_method_importance(
        "dependent_features", n, response_type, seed, "correlation",
        results_dir
    ) |>
        rename(correlation = importance)
    kn <- load_method_importance(
        "dependent_features", n, response_type, seed, "knockoffs",
        results_dir
    ) |>
        rename(knockoffs = importance)

    corr |>
        select(feature, correlation) |>
        left_join(select(kn, feature, knockoffs), by = "feature") |>
        filter(feature %in% c("x1", "x2")) |>
        mutate(seed = seed, n = n, response_type = response_type)
})
```

```{r fig-divergence-corr, fig.width=7.5, fig.height=3}
divergence_means_corr <- divergence_df_corr |>
    group_by(feature, response_type, n) |>
    summarize(
        correlation = mean(correlation), knockoffs = mean(knockoffs),
        .groups = "drop"
    )

p_div_corr <- ggplot(
    divergence_df_corr,
    aes(correlation, knockoffs, color = feature, shape = response_type)
) +
    geom_hline(yintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_vline(xintercept = 0, color = axiom_palette$grid, linewidth = 1.5) +
    geom_point(size = 2, alpha = 0.4) +
    geom_point(data = divergence_means_corr, size = 3, alpha = 0.8) +
    facet_wrap(~n, labeller = labeller(n = ~ glue("n = {.x}"))) +
    scale_color_manual(
        values = feature_colors, labels = feature_labels, name = NULL
    ) +
    scale_shape_manual(values = c(classification = 16, regression = 17)) +
    labs(
        title = glue("Correlation vs. knockoffs on dependent_features"),
        x = "Correlation importance", y = "Knockoffs importance",
        shape = "Response type"
    )

p_div_corr
ggsave(
    path(fig_dir, "fig_divergence_correlated.svg"), p_div_corr,
    width = 7.5, height = 3
)
```

# minSHAP mediated-chain diagnostic

The synthetic mediated-chain case study has two chains,
$x_1 \to x_2 \to x_3 \to Y$ and
$x_4 \to x_5 \to x_6 \to Y$. Only $x_3$ and $x_6$ are conditionally related
to the response after all features are observed. The plots below use the exact
sampled orderings and predecessor coalitions that produced the reported
minSHAP and SAGE results. They do not enumerate a separate coalition game.

```{r load-mediated-chain-contributions}
mediated_stem <- with(
    mediated_focus,
    glue("mediated_chains_{n}_{response_type}_{seed}")
)
mediated <- load_mediated_contributions(mediated_stem, results_dir)
mediated_contributions <- mediated$contributions
mediated_membership <- mediated$membership
mediated_aggregates <- mediated$aggregates
```

## Exact sampled conditional contributions

```{r fig-mediated-contribution-focus, fig.width=15, fig.height=18}
focus <- mediated_focus_data(mediated_membership, mediated_contributions, FEATURE_ORDER)
p_mediated_focus <- mediated_membership_plot(
    focus$membership, focus$outline,
    subtitle = "Outline: added feature j"
) | mediated_bars_plot(focus$bars)
p_mediated_focus

ggsave(
    path(fig_dir, "fig_mediated_chain_contribution_focus.svg"),
    p_mediated_focus, width = 15, height = 18
)
```

```{r fig-mediated-contribution-highlight, fig.width=10, fig.height=5}
mediation_highlight <- c("noise_1", "x1", "x3")
highlight <- mediated_focus_data(
    mediated_membership, mediated_contributions, mediation_highlight
)
p_mediated_highlight <- mediated_membership_plot(
    highlight$membership, highlight$outline,
    title = "Coalitions",
) | mediated_bars_plot(highlight$bars)
p_mediated_highlight

ggsave(
    path(fig_dir, "fig_mediated_chain_contribution_highlight.svg"),
    p_mediated_highlight, width = 10, height = 5
)
```

## Aggregated contributions

```{r fig-mediated-aggregate-hinton}
#| fig-width: 10
#| fig-height: 8
p_mediated_hinton <- mediated_hinton_plot(mediated_aggregates)
p_mediated_hinton

ggsave(
    path(fig_dir, "fig_mediated_chain_aggregate_hinton.svg"),
    p_mediated_hinton, width = 12, height = 4.5
)
```
